---
title: "Canada's Population Change in Shiny"
author: "Kyle Rozic"
date: "4/22/2021"
output: slidy_presentation
---

### Introduction

My goal for this project was to work more with maps to create choropleth visualizations of Canadian population. All data was obtained from publicly available datasets at Stats Canada.

My final application allows the user to select years from a drop down to visualize the population change per 10k citizens on a leaflet choropleth map. The only library that may not be recognized by classmates is `sf` which is used to handle shape data (for maps).

```{r, warning = F, message = F}
library(shiny)
library(dplyr)
library(data.table)
library(sf)
library(RColorBrewer)
library(leaflet)
library(tidyr)
library(ggplot2)
```

### The Data

After spending a lot of time finding and preparing the data to work with this project, I made 2 final datasets that are used for this final project: A tall dataset containing population values for each census district factorized by year, and the shape data to draw the cartographic boundaries.

```{r}
popTot <- read.csv('./data/popTot.csv', 
                   na.strings = '')
shp <- st_read('./data/division_shapes_digital/lcd_000a16a_e.shp', 
               stringsAsFactors = F)
head(popTot)
head(shp)
```

### Visualizing with `ggplot2`

In R, districts are quickly visualized with `ggplot2`. First I spread the data so it's more easily handled, then I merged it with the shape data:

```{r}
popTot <- spread(popTot, YEAR, POPULATION) %>%
        select(-GEO)
popMap <- st_as_sf(merge(popTot, shp))

ggplot() + 
        geom_sf(data = popMap, aes(fill = (`2020`-`2019`)/`2019`*10000)) +
        scale_fill_gradient2()
```

### Notes on shiny implementation

I thought the shiny application would be a pretty straight forward implementation of the above `ggplot2` code, however the reactive environment introduced some troubles which were difficult to troubleshoot in the reactive environment. 

For example, reactive elements cannot be transformed on the right side of the assignment operation and reassigned to itself on the left side, as this will create an infinite loop of reactivity. (ex. `val <- val + 1`)

Also, while subsetting reactive data frames does behave normally most of the time, I did notice it break once or twice. Sometimes calling a single column or variable from a data frame returned a list containing a vector, rather than just the vector itself. This caused problems with calculations that were working before. 

Anyway, here is the code:

```{r, eval = F}
library(shiny)
library(shinydashboard)
library(leaflet)

shinyUI(dashboardPage(
    skin = 'blue',
    dashboardHeader(title = 'Canada Population Change'),
    dashboardSidebar(
        h1('Choose Year'),
        selectInput('date1', 'Start Date:', seq(2001, 2020)),
        selectInput('date2', 'End Date:', seq(2001, 2020), selected = 2002),
        h1('Choose Table Type'),
        selectInput('tableType', 'Contents:', c('All Years', 'Change')),
        column(12, 
               style = 'margin-left: 70px', 
               submitButton('Apply'))
    ),
    dashboardBody(
        fluidRow(box(width = 12, leafletOutput(outputId = 'map', height = 500))),
        fluidRow(box(width = 12, dataTableOutput(outputId = 'summary_table')))
    )
))
```

```{r, eval = F}
library(shiny)
library(dplyr)
library(data.table)
library(sf)
library(RColorBrewer)
library(leaflet)
library(tidyr)

popTot <- read.csv('./data/popTot.csv', 
                   na.strings = '')
popTot <- spread(popTot, YEAR, POPULATION) %>%
        select(-GEO)
shp <- st_read('./data/division_shapes_digital/lcd_000a16a_e.shp', 
               stringsAsFactors = F)

shinyServer(function(input, output) {
        date1 <- reactive({grep(input$date1, names(popTot))})
        date2 <- reactive({grep(input$date2, names(popTot))})
        inputData <- reactive({
                popTot %>%
                        mutate(CHANGE_PER_10k = round((popTot[date2()]-popTot[date1()])/popTot[date1()]*10000)) %>%
                        select(CITY, PROVINCE, CHANGE_PER_10k, CDUID)
        })
        popMap <- reactive({st_as_sf(merge(inputData(), shp)) %>%
                st_transform('+init=epsg:4326')})

        minVal <- reactive({min(popMap()$CHANGE_PER_10k)})
        maxVal <- reactive({max(popMap()$CHANGE_PER_10k)})
        domain <- reactive({c(minVal(), maxVal())})
        colorPal <- reactive({c(colorRampPalette(colors = brewer.pal(11, 'RdBu')[c(1:4, 6)],
                                       space = 'Lab')(abs(minVal()/10)),
                      colorRampPalette(colors = brewer.pal(11, 'RdBu')[c(6, 8:11)],
                                       space = 'Lab')(maxVal()/10)[-1])})
        colorPalInput <- reactive({colorNumeric(colorPal(), domain())})

        labels = reactive({sprintf('<strong>%s, %s</strong><br/>%g net migration/10k',
                         popMap()$CITY,
                         popMap()$PROVINCE,
                         popMap()$CHANGE_PER_10k[[1]]) %>%
                lapply(htmltools::HTML)})

        output$map <- renderLeaflet({
                popMap() %>%
                leaflet() %>%
                addProviderTiles('Stamen.TonerLite') %>%
                setView(lng = -95, lat = 60, zoom = 3) %>%
                addPolygons(color = '#444444',
                            weight = 1,
                            smoothFactor = 0.5,
                            opacity = 1,
                            fillOpacity = 0.7,
                            fillColor = ~colorPalInput()(popMap()$CHANGE_PER_10k[[1]]),
                            highlightOptions = highlightOptions(color = 'white',
                                                                weight = 2,
                                                                bringToFront = T),
                            label = labels(),
                            labelOptions = labelOptions(
                                    style = list('font-weight' = 'normal',
                                                 padding = '3px 8px'),
                                    textsize = '15px',
                                    direction = 'auto')) %>%
                addLegend(pal = colorPalInput(),
                          values = domain(),
                          opacity = 0.7,
                          title = 'Net Migration Per 10k',
                          position = 'bottomright')
        })
        tableType <- reactive({if(input$tableType == 'All Years'){popTot} else{inputData()}})
        output$summary_table <- renderDataTable(tableType(), 
                                                options = list(pageLength = 25,
                                                               width = '100%',
                                                               scrollx = T))
})
```

I used `shinydashboard` to make a slightly different looking web UI. 

[shiny link](https://krozic.shinyapps.io/shiny_app/)

To use the app:

1. Choose the dates you would like to observe the change for.
2. Choose the type of table you would like to view.
3. Click 'Apply'.

**Have a great day stranger!**